# Rarefying

## About Rarefying

Rarefying is another common normalization technique that standardizes the library size across samples that was originally used in ecology. This method standardizes the read depth across all samples. To perform this method we first choose a minimum library size. Looking at rarefaction/collectors curves, or using a certain percentile can guide choosing this cutoff. Then all samples that have a read depth below this cutoff are discarded. Thus this method has a built-in filtering step. Next, we sample without replacement of the size of the chosen cutoff. It can be a standalone method or combined with other methods and transformations.

<!-- Fix second to last sentence - confusing  -->

This is a very commonly used method, but it has also been criticized [@mcmurdie2014]. First of all, it throws away valid data, and this results in a loss of power and an increase in false positives. Rare taxa can be removed in this approach too. It is however encouraged when we have widely different library sizes as it can lower the false discovery rate [@weiss2017], and has also been shown to perform well in community-level analysis [@mcknight2019], as it completely standardizes the read count depth, and some methods are sensitive to differences in read count. Rarefying has been shown to separate by biological signal in ordination methods based on presence/absence.

## Rarefying implementation 

Consider the following example dataset. 


```{r, echo = F}
sample1 <- c(132, 7, 0, 23, 71)
sample2 <- c(103, 48, 2, 15, 80)
sample3 <- c(11, 3, 1, 2, 5)
taxa <- paste("Taxon", 1:5)

sampling_depth <- c(sum(sample1), sum(sample2), sum(sample3))

tibble(taxa, sample1, sample2, sample3) %>%
    knitr::kable(
        col.names = c("Taxa", "Sample 1", "Sample2", "Sample 3"),
        caption =  "Example dataset - Raw Counts", 
        format = "html"
    ) 

# %>% kableExtra::row_spec(5, extra_css = "border-bottom: 1px solid")
```

To normalize this data by rarefying, we first choose a minimum sampling depth, or in other words, the minimum column sum. After normalizing, the column sum of all samples will be this size. Any samples that are below this minimum will be dropped from analysis. We choose 200 as our minimum here. In practice the counts will normally be much higher. In this example, our sampling depths are 223, 248, and 20. Thus the third sample will be dropped in this procedure. 

Since we are randomly sampling, we need to keep track of the seed so this process is reproducible. 

```{r}
set.seed(525)
```

We then randomly sample taxa from each sample/column in the proportions corresponding to the raw counts. The below table shows the table from above after we rarefy. Notice that the rare Taxon 3, which originally was present in Sample 1 and 3 is now completely removed from the analysis. 

```{r, echo = F}

# r1 <- data.frame(taxa, sample1) %>% 
#     uncount() %>% 
#     dplyr::pull(taxa) %>% 
#     sample(size = 200 ) %>% table() %>% as.vector(
#     )
# 
# r2 <- data.frame(taxa, sample2) %>% 
#     uncount() %>% 
#     dplyr::pull(taxa) %>% 
#     sample(size = 200) %>% table() %>% as.vector()
# r1
# r2
r1 <- c(115, 6, 0, 18, 61)
r2 <- c(90, 39, 0, 11, 60)

tibble(taxa, r1, r2) %>%
    knitr::kable(col.names = c("Taxa", "Sample 1", "Sample2"),
        caption =  "Rarefied counts") 
```




We now show the process of rarefying a real dataset. The following function returns a rarefied `phyloseq` object. We can either pass in the minimum sampling depth as a second argument, or use the default minimum depth of the samples. 

```{r, gp_rare}
norm_rarefy <- function(phyloseq, depth = min(sample_sums(phyloseq))){
    return(phyloseq::rarefy_even_depth(phyloseq,sample.size = depth))
}
    
```

## Rarefying on Global Patterns

We use the above function to rarefy the Global Patterns data. The first difficulty is choosing a minimum sampling depth. The Global Patterns dataset already has a very high sampling depth for all samples, so we will chose the lowest as the minimum depth to rarefy to.Since we chose the minimum sampling depth, no samples have been dropped. In data sets where we have low sampling depth there is a balance between how many samples to drop and how low to set the minimum depth to.

```{r}
gp_rare <- norm_rarefy(gp_raw)
```

We can check that indeed all samples now have the same sampling depth, which is 15905. Note that the highest sampling depth in this dataset was almost 2 million, so we have discarded a lot of data to reduce to 15905.

```{r sample_sums_after_rare}
max(sample_sums(gp_raw))
sample_sums(gp_rare)
```

We can again compare the PCoA plots between rarefied and raw counts, coloring by sample type to view clusters. 

```{r, rare_ordination}
plot_ordination(gp_raw,
                phyloseq::ordinate(gp_raw, "PCoA", "bray"),
                color = "SampleType", 
                title = "PCoA on Raw counts") +
plot_ordination(gp_rare,
                phyloseq::ordinate(gp_rare, "PCoA", "bray"),
                color = "SampleType", 
                title = "PCoA on Rarefied counts") +
  plot_layout(guides = 'collect')
```


Now examine how the distance matrices change before/after normalization. We see a similar pattern to TSS when distance matrices calculated from rarefied counts are compared to those from the raw counts. 

```{r, rare_plot_ordination}
# Identify any samples filtered in rarefying process 
rare_samples <- sample_names(gp_rare)
gp_raw_match <- prune_samples(rare_samples, gp_raw)
plot_norm_changes(gp_rare, gp_raw_match,
                  x_lab = "Raw counts", y_lab = "Rarefied counts", 
                  title = "Distance metric comparision between Raw counts and Rarefied counts ")

## Compare to tss
gp_tss_match <- norm_TSS(gp_raw_match)
plot_norm_changes(gp_rare, gp_tss,
                  x_lab = "TSS", y_lab = "Rarefied", 
                  title = "Distance metric comparision between TSS normalization and Rarefied counts ")

```



## Rarefying on Kostic Data

```{r}
k_rare <- norm_rarefy(k_raw)

plot_norm_changes(k_rare, k_raw,
                  x_lab = "Raw counts", y_lab = "Rarefied counts", 
                  title = "Distance metric comparision between Raw counts and Rarefied counts ")

plot_norm_changes(k_rare, k_tss,
                  x_lab = "TSS", y_lab = "Rarefied", 
                  title = "Distance metric comparision between TSS normalization and Rarefied counts ")
```


<!-- Doesnt seem to be much difference between rarefied and tss -->

<!-- ## Rarefying on HMP dataset -->

<!-- ```{r, eval = F} -->

<!-- summary(sample_sums(hmp_raw)) -->

<!-- hmp_rare <- rarefy_even_depth(hmp_raw, -->

<!--                              sample.size = 3000) -->

<!-- ``` -->

<!-- Ordination  -->

<!-- ```{r, eval = F} -->

<!-- plot_ordination(hmp_raw, -->

<!--                 hmp_raw_dist, color = "sampling_depth",  -->

<!--                 title = "Raw counts") -->

<!-- hmp_rare_dist <- phyloseq::ordinate(hmp_rare, "PCoA", "bray") -->

<!-- plot_ordination(hmp_rare, -->

<!--                 hmp_rare_dist, -->

<!--                 color = "sampling_depth",  -->

<!--                 title = "Rarefied counts") -->

<!-- ``` -->

<!-- Compare distances - need to work on this since with rarefying there are fewer points -->

<!-- ```{r} -->

<!-- #plot_norm_changes(hmp_rare, hmp_tss) -->

<!-- ``` -->
